Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2023-11-09 Journal-manuscript/Journal-manuscript.tex
Antonio completed editions of revised version of paper; now it is going back to Anastasia for proof-reading

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
from the perspective of planning too {[}see, \emph{inter alia},
4,5--8{]}. Beginning with the work of Hansen {[}1{]}, accessibility
measures have been widely used to evaluate the efficiency of
transportation systems when combined with the distribution of
opportunities in space {[}e.g., {]}. As such, it is a holistic measure
of spatial systems that measures the ease of reaching destinations
{[}9,10{]}.

@@

The issue of congestion was the motivation for the development of
floating catchment area approaches in accessibility analysis
{[}11,12{]}. While these approaches purport to account for congestions,
Paez et al. {[}13{]} demonstrate that in general they do not solve the
issue of multiple counting of opportunities, thus leading to biases in
the calculation of total demand and supply, sometimes inflating them,
other times deflating them. In response to this, recent research has
paid closer attention to the way opportunities are counted in
accessibility analysis. {[}13{]}, for example, tackle floating catchment
area methods and introduce a normalization of the impedance matrix to
allocate the population and then the level of service proportionally.
More recently, Soukhov et al. {[}14{]} introduced a singly-constrained
measure of accessibility, called spatial availability, that employs a
similar, but more sophisticated proportional allocation mechanism. The
work of these authors shows that floating catchment area methods can be
seen as singly-constrained accessibility measures, and improve on
existing approaches by guaranteeing that each opportunity is counted
only once - in other words, treating opportunities as \emph{finite}. The
proportional allocation of spatial availability constrains the
calculations to match a known quantity, therefore ensuring that the
measurements sum up to a predetermined quantity (i.e., the total number
of opportunities), and so each value can be meaningfully related to this
total.

A limitation of spatial availability as introduced by Soukhov et al.
{[}14{]} is that it was developed for the case of a homogeneous
population, for example for the case of a single mode of transportation.
However, the finite nature of opportunities makes the analysis of
heterogeneous populations very relevant. In the case of multiple modes
of transportation, people who travel by slow modes (e.g., active modes)
can usually reach fewer opportunities than people who travel by faster
modes and whose range is typically far wider (e.g., car). This implies
that slower travelers will often face increased competition for local
opportunities from travelers who can reach said opportunities from
farther afield.

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but it is worthwhile noting
that this is in fact just one case of heterogeneous populations (i.e.,
travel by different modes). The method itself can easily accommodate
other forms of heterogeneity, for example variations in travel behavior
between older and younger adults {[}e.g., 15{]}, the propensity of older
adults to use different modes of transportation {[}16{]}, the usually
shorter trip lengths of children compared to grown-ups {[}e.g., 17{]},
or the more limited travel ranges of single parents {[}e.g., 18{]}.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}19{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}20{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}20{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analyzed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}21{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'' (e.g., in the abstract), it is unclear what the
unconstrained scores mean: is having access to but 10,000 jobs by
transit insufficient? After all, 10,000 employment opportunities are
still plenty of opportunities. These ratios can be found elsewhere in
the literature {[}e.g., 22, and 9,15,18, and 8{]}, and they are useful
to assess when some members of the public are better or worse off than
others, but they do not say much about how bad is ``worse''.

Besides ratios of accessibility, another way to improve interpretability
of scores sometimes seen in the literature is to standardize them to lie
in the range {[}0-1{]}. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardized in this way is not
uncommon in the literature {[}e.g., 23,24{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}11,14,25--29{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are clearly competitive.
But we would go as far as to argue that every type of opportunity is
subject to congestion or capacity constraint, even when the
opportunities are conventionally seen as non-competitive. Amenities are
a good example of this. For instance, standards for providing green
spaces are often stated in the form of \emph{exclusive access}, in units
of amenity per capita. A case in point is Natural England, an
organization that recommends an Accessible Natural Greenspace Standard
such that the minimum supply of space is one ha of statutory Local
Nature Reserves per thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 30{]} recommends
that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of a
resource can be offered to each individual that is not allocated to
anyone else?''. Green spaces often have large capacities, but they still
have a capacity. Constraining accessibility is in this way a useful way
to evaluate the congested availability of any type of opportunity. As
development of sound standards is emphasized in the planning literature,
in particular in regards to fairness in transportation {[}see 31{]},
spatial availability analysis is a useful way to develop and assess
standards.

The relevance of the considerations above is put in sharper relief when
we think about the use of multiple modes (or heterogeneous populations).
If we return to Oslo for a moment {[}21{]}, we notice that the places
that have high accessibility by transit are also the places that have
\emph{very high} accessibility by car (Figure 2). Those two populations
are going for the same opportunities, and those travelling by transit
have fewer to choose from to begin with. More generally, people in a
zone who are advantaged with relatively cost of travel will have the
ability to potentially reach more opportunities than other people. Due
to this advantage, through the perspective of finite opportunities,
there are fewer opportunities left for everyone else, especially for
those who use modes that are slower or more expensive.

As noted in the introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by {[}12,11,although similar, and earlier, developments are
found in 29,32{]}. Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 13{]}. In other words, interpretation of differences in Shen-type
accessibility scores between modes is fraught as it is for Hansen-type
measures.

To illustrate, {[}33{]} calculate \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on previously discussed accessibility
approaches using the Hansen-type measure and the Shen-type measure by
constraining the sum of opportunities, that is, by treating
opportunities as finite. This is done by means of proportional
allocation factors that follow well established principles of spatial
interaction and the gravity model {[}see 34{]}. In Soukhov et al.
{[}14{]} these factors consider: 1) the mass effect (e.g., the size of
populations at different origins); and 2) the cost of travel from
different zones (e.g., some sub-populations face relatively higher or
lower costs). The following section introduces the multimodal form of
spatial availability.

\hypertarget{multimodal-spatial-availability-v_im}{%
\section{\texorpdfstring{Multimodal spatial availability
\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}

In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
proportion of all opportunities in the region \(O\) that are allocated
to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
a value of how many opportunities are available to each location \(i\)
out of all the opportunities in the region. The general formulation of
spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}) {[}see 14{]}:

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population, followed by then \(A\), then \(C\). This synthetic example
is inspired by the single-mode example used in {[}11{]} and reconfigured
in {[}14{]}.

@@

{\centering \includegraphics[width=1\linewidth]{images/Fig1}

@@

The values calculated for \(S_i^m\) (Hansen-type accessibility),
\(a_i^m\) (Shen-type accessibility), and \(V_i^m\) (spatial
availability) for each \(i\) and \(m\) are shown in the middle three
columns and are aggregated for each \(i\) in the final two columns in
Table 1 . As in the example in Shen {[}11{]}, we use a negative
exponential impedance function
\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.
Notice that in this example we use the same impedance function but the
travel times are different for the two modes. More generally, it is
possible to use different impedance functions for the modes, as
demonstrated in the empirical example below.

Hansen-type accessibility \(S_i^m\) is presented for each origin and
mode in the third column of Table 1 . For all \(i\), the travel by \(z\)
results in higher values of \(S_i^m\) than travel by \(x\). Lack of
competition, or alternatively the assumption of an inexhaustible
resource in the calculation of \(S_i^m\), lead to a curious result.
Since the populations in \(A\) and \(B\) have the same travel impedance
to employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100
minutes using \(x\) or 10, 25, or 80 minutes using \(z\)), their values
of \(S_i^m\) are the same for both \(A\) and \(B\). Furthermore, the
total sum of \(S_i^m\) in the region is equal to 150,570.2. This value
lacks an intuitive interpretation: it represents the weighted sum of
opportunities that may be reached within the region according to the
travel impedance (i.e., the travel behavior and the characteristics of
the modes) and does not usefully translate into any sort of benchmark.
To connect this example to the literature, \(S_i^m\) is calculated in
the work of {[}20{]}; they compare differences in \(S_i^m\) values
between modes in a relative and comparative sense, but make no further
interpretation of the \(S_i^m\) values. Large metropolitan regions will
have large values and small regions small, but that is simply an
artifact of size.

@@

However, the calculations of \(a_i^m\) are not constrained: the total
sum of \(a_i^m\) or \(a_i\) is practically meaningless since it
represents a sum of ratios. For instance, the population travelling by
\(z\) from \(A\) has a value of 1.57 jobs per job-seeking population
compared to 0.95 for users of mode \(x\). What is the meaning of these
values? The difference between these modes is equal to 0.62, but 0.62 of
what? How many more job opportunities can users of \(z\) reach compared
to user of \(x\)? When \(a_i^m\) is aggregated to \(a_i\) as shown in
the sixth column, the values face similar interpretability issues. The
Shen-type measure is implemented in aforementioned work of Tao et al.
{[}33{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
is implemented in the work of {[}35{]}. However, similar to Hansen-type
accessibility, these works discuss relative and spatially comparative
differences in values, but veer from interpretating the values of
\(a_i^m\) or \(a_i\) themselves. In fairness, interpretation is
complicated by the multiple counting of opportunities between zones and
modes.

@@

\hypertarget{empirical-example-madrid-lez}{%
\section{Empirical example: Madrid
LEZ}\label{empirical-example-madrid-lez}}

\hypertarget{data-and-methods}{%
\subsection{Data and methods}\label{data-and-methods}}

The context for the empirical example is Madrid, in Spain. This city
implemented a Low Emission Zone in 2017, to attend to goals set out in
the national agenda to fight climate change, cut nitrogen dioxide
levels, and prioritize people's movement in the city. Low emission zones
(LEZ) elsewhere have been implemented as a climate change policy
intervention to reduce GHG emissions, improve air quality, and support
sustainable mobility {[}36,37{]}. Though the rules of exclusion vary,
LEZ aim to deter/reduce traffic in designated zones under threat of
penalty (e.g., fines, seizure of vehicle). In other words, LEZ are a
policy of \emph{geographic discrimination} as they change how people can
reach opportunities by making the travel impedance more costly for
specific forms of travel, typically cars. When considering opportunities
as finite in a region, this discrimination reduces the competition of
one mode and allows other modes to potentially thrive. In this way, LEZ
aim to change the multimodal competitive accessibility landscape of a
city.

Spain is one of a few countries with active LEZ and plans to expand
their implementation as specified in their climate-change-related plans:
\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}38{]} and
\emph{Plan Nacional de Control de la Contaminación Atmosférica}
{[}39{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
Cambio Climático y Transición Energética}) will require all
municipalities to implement LEZ by 2023 if they meet at least one of the
following requirements: (i) municipalities \textgreater50,000 inhab.;
(ii) islands; and (iii) municipalities \textgreater{} 20,000 inhab. when
air quality exceeds limits specified in \emph{RD 102/2011 de Mejora de
Calidad del Aire} {[}40{]}.

In geographic scope, the 2017 boundaries of the LEZ in Madrid were
relatively modest (covering 4.72 km\(^2\)) and only within the center
(i.e., LEZ Centro). These boundaries were expanded in 2023 to the area
inside the M-30, an orbital highway in proximity to the city center
(i.e., LEZ M-30). Beyond this, the city has plans to spatially expand
the LEZ. Within the 2017 LEZ Centro implementation, all cars,
motorcycles and freight with environmental labels A or B (older makes
and models of fossil fuel internal combustion engine vehicles), are not
permitted to enter the area unless they are used by residents or meet
other exemptions. This restriction impacted approximately half of all
car trips that typically travelled into what is now the LEZ Centro
{[}41{]}.

>>>>

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
from the perspective of planning too {[}see, \emph{inter alia},
4,5--8{]}. Beginning with the work of Hansen {[}1{]}, accessibility
measures have been widely used to evaluate the efficiency of
transportation systems when combined with the distribution of
opportunities in space {[}9{]}. As such, it is a holistic measure of
spatial systems that measures the ease of reaching destinations
{[}10,11{]}.

@@

The issue of congestion was the motivation for the development of
floating catchment area approaches in accessibility analysis
{[}12,13{]}. While these approaches purport to account for congestions,
Paez et al. {[}14{]} demonstrate that in general they do not solve the
issue of multiple counting of opportunities, thus leading to biases in
the calculation of total demand and supply, sometimes inflating them,
other times deflating them. In response to this, recent research has
paid closer attention to the way opportunities are counted in
accessibility analysis. {[}14{]}, for example, tackle floating catchment
area methods and introduce a normalization of the impedance matrix to
allocate the population and then the level of service proportionally.
More recently, Soukhov et al. {[}15{]} introduced a singly-constrained
measure of accessibility, called spatial availability, that employs a
similar, but more sophisticated proportional allocation mechanism. The
work of these authors shows that floating catchment area methods can be
seen as singly-constrained accessibility measures, and improve on
existing approaches by guaranteeing that each opportunity is counted
only once - in other words, treating opportunities as \emph{finite}. The
proportional allocation of spatial availability constrains the
calculations to match a known quantity, therefore ensuring that the
measurements sum up to a predetermined quantity (i.e., the total number
of opportunities), and so each value can be meaningfully related to this
total.

A limitation of spatial availability as introduced by Soukhov et al.
{[}15{]} is that it was developed for the case of a homogeneous
population, for example for the case of a single mode of transportation.
However, the finite nature of opportunities makes the analysis of
heterogeneous populations very relevant. In the case of multiple modes
of transportation, people who travel by slow modes (e.g., active modes)
can usually reach fewer opportunities than people who travel by faster
modes and whose range is typically far wider (e.g., car). This implies
that slower travelers will often face increased competition for local
opportunities from travelers who can reach said opportunities from
farther afield.

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but it is worthwhile noting
that this is in fact just one case of heterogeneous populations (i.e.,
travel by different modes). The method itself can easily accommodate
other forms of heterogeneity, for example variations in travel behavior
between older and younger adults {[}e.g., 16{]}, the propensity of older
adults to use different modes of transportation {[}17{]}, the usually
shorter trip lengths of children compared to grown-ups {[}e.g., 18{]},
or the more limited travel ranges of single parents {[}e.g., 19{]}.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}20{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}21{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}21{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analyzed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}22{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'' (e.g., in the abstract), it is unclear what the
unconstrained scores mean: is having access to but 10,000 jobs by
transit insufficient? After all, 10,000 employment opportunities are
still plenty of opportunities. These ratios can be found elsewhere in
the literature {[}e.g., 23, and 9,16,19, and 8{]}, and they are useful
to assess when some members of the public are better or worse off than
others, but they do not say much about how bad is ``worse''.

Besides ratios of accessibility, another way to improve interpretability
of scores sometimes seen in the literature is to standardize them to lie
in the range {[}0-1{]}. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardized in this way is not
uncommon in the literature {[}e.g., 24,25{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}12,15,26--30{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are unambiguously
competitive. But we would go as far as to argue that every type of
opportunity is subject to congestion or capacity constraint, even when
the opportunities are conventionally seen as non-competitive. Amenities
are a good example of this. For instance, standards for providing green
spaces are often stated in the form of \emph{exclusive access}, in units
of amenity per capita. A case in point is the Ile-de-France region, a
jurisdiction that suggested in a major planning document of 2013 that at
the municipal level at least \(10m^2\) of public green space should be
supplied \emph{per inhabitant} {[}31{]}. But green spaces are not evenly
distributed, which means that who has access to them hinges on where
they are and how easy is to reach them. This formulation of provision of
amenities is not unusual. For example, Natural England, an organization
that recommends an Accessible Natural Greenspace Standard such that the
minimum supply of space is one ha of statutory Local Nature Reserves per
thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 32{]} recommends
that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of
this resource is available to one individual that has not been claimed
by anyone else?''. Green spaces often have large capacities, but they
still have a capacity, and it is not the same a person to have access to
\(5m^2\) of \emph{uncongested} green space than to \(15m^2\). This
difference is in fact a matter of justice {[}31,33{]}. Constraining
accessibility is in this way a useful way to evaluate the congested
availability of any type of opportunity. As development of sound
standards is emphasized in the planning literature, in particular in
regards to fairness in transportation {[}see 34{]}, spatial availability
analysis is a useful way to develop and assess standards.

The relevance of the considerations above is put in sharper relief when
we think about the use of multiple modes (or heterogeneous populations).
If we return to Oslo for a moment {[}22{]}, we notice that the places
that have high accessibility by transit are also the places that have
\emph{very high} accessibility by car (Figure 2). Those two populations
are going for the same opportunities, and those travelling by transit
have fewer to choose from to begin with. More generally, people in a
zone who are advantaged with relatively cost of travel will have the
ability to potentially reach more opportunities than other people. Due
to this advantage, through the perspective of finite opportunities,
there are fewer opportunities left for everyone else, especially for
those who use modes that are slower or more expensive.

As noted in the introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by {[}13,12,although similar, and earlier, developments are
found in 30,35{]}. Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 14{]}. In other words, interpretation of differences in Shen-type
accessibility scores between modes is fraught as it is for Hansen-type
measures.

To illustrate, {[}36{]} calculate \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on previously discussed accessibility
approaches using the Hansen-type measure and the Shen-type measure by
constraining the sum of opportunities, that is, by treating
opportunities as finite. This is done by means of proportional
allocation factors that follow well established principles of spatial
interaction and the gravity model {[}see 37{]}. In Soukhov et al.
{[}15{]} these factors consider: 1) the mass effect (e.g., the size of
populations at different origins); and 2) the cost of travel from
different zones (e.g., some sub-populations face relatively higher or
lower costs). The following section introduces the multimodal form of
spatial availability.

\hypertarget{multimodal-spatial-availability}{%
\section{Multimodal spatial
availability}\label{multimodal-spatial-availability}}

In brief, we define the spatial availability at an origin \(i\)
(\(V_{i}\)) as the proportion of all opportunities in the region that
are allocated to location \(i\) from all destinations \(j\). \(V_{i}\)
is a value of how many opportunities are available to each location
\(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}) {[}see 15{]}:

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population, followed by then \(A\), then \(C\). This synthetic example
is inspired by the single-mode example used in {[}12{]} and reconfigured
in {[}15{]}.

@@

{\centering \includegraphics[width=0.85\linewidth]{images/Fig1}

@@

The values calculated for \(S_i^m\) (Hansen-type accessibility),
\(a_i^m\) (Shen-type accessibility), and \(V_i^m\) (spatial
availability) for each \(i\) and \(m\) are shown in the middle three
columns and are aggregated for each \(i\) in the final two columns in
Table 1 . As in the example in Shen {[}12{]}, we use a negative
exponential impedance function
\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.
Notice that in this example we use the same impedance function but the
travel times are different for the two modes. More generally, it is
possible to use different impedance functions for the modes, as
demonstrated in the empirical example below.

Hansen-type accessibility \(S_i^m\) is presented for each origin and
mode in the third column of Table 1 . For all \(i\), the travel by \(z\)
results in higher values of \(S_i^m\) than travel by \(x\). Lack of
competition, or alternatively the assumption of an inexhaustible
resource in the calculation of \(S_i^m\), lead to a curious result.
Since the populations in \(A\) and \(B\) have the same travel impedance
to employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100
minutes using \(x\) or 10, 25, or 80 minutes using \(z\)), their values
of \(S_i^m\) are the same for both \(A\) and \(B\). Furthermore, the
total sum of \(S_i^m\) in the region is equal to 150,570.2. This value
lacks an intuitive interpretation: it represents the weighted sum of
opportunities that may be reached within the region according to the
travel impedance (i.e., the travel behavior and the characteristics of
the modes) and does not usefully translate into any sort of benchmark.
To connect this example to the literature, \(S_i^m\) is calculated in
the work of {[}21{]}; they compare differences in \(S_i^m\) values
between modes in a relative and comparative sense, but make no further
interpretation of the \(S_i^m\) values. Large metropolitan regions will
have large values and small regions small, but that is simply an
artifact of size.

@@

However, the calculations of \(a_i^m\) are not constrained: the total
sum of \(a_i^m\) or \(a_i\) is practically meaningless since it
represents a sum of ratios. For instance, the population travelling by
\(z\) from \(A\) has a value of 1.57 jobs per job-seeking population
compared to 0.95 for users of mode \(x\). What is the meaning of these
values? The difference between these modes is equal to 0.62, but 0.62 of
what? How many more job opportunities can users of \(z\) reach compared
to user of \(x\)? When \(a_i^m\) is aggregated to \(a_i\) as shown in
the sixth column, the values face similar interpretability issues. The
Shen-type measure is implemented in aforementioned work of Tao et al.
{[}36{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
is implemented in the work of {[}38{]}. However, similar to Hansen-type
accessibility, these works discuss relative and spatially comparative
differences in values, but veer from interpretating the values of
\(a_i^m\) or \(a_i\) themselves. In fairness, interpretation is
complicated by the multiple counting of opportunities between zones and
modes.

@@

\hypertarget{empirical-example}{%
\section{Empirical example}\label{empirical-example}}

\hypertarget{context}{%
\subsection{Context}\label{context}}

The context for the empirical example is Madrid, Spain. This city
implemented a Low Emission Zone in 2017, to pursue goals set out in the
national climate change agenda, to cut nitrogen dioxide levels, and to
prioritize people's movement in the city. Low Emission Zones (LEZs)
elsewhere have been implemented as interventions to reduce GHG
emissions, improve air quality, and support sustainable mobility
{[}39,40{]}. Though the rules of exclusion vary by city, LEZs aim to
deter/reduce traffic in designated zones under threat of penalty (e.g.,
fines, seizure of vehicle). In other words, LEZ implement a form of
\emph{geographic discrimination} as they change how people can reach
opportunities by making it more costly for some forms of travel,
typically cars, to circulate in predetermined zones. When considering
opportunities as finite in a region, this discrimination reduces the
competition of one mode and opens up opportunties for other modes to
thrive. In this way, at their core LEZs operate by changing the
accessibility landscape of a city from the perspective of multiple
modes.

In geographic scope, the 2017 boundaries of the LEZ in Madrid were
relatively modest, covering only approximately 4.72 \(km^2\) of the
central business district of the city (the so-called LEZ Centro). As of
this writing, there are plans to expand these boundaries to the area
inside the M-30, an orbital highway in proximity to the city center
(i.e., LEZ M-30). Within the 2017 LEZ Centro implementation, all cars,
motorcycles and freight vehicles with environmental labels A or B (older
makes and models of fossil fuel internal combustion engine vehicles),
were disallowed from entering the zone unless they are used by residents
or meet other exemptions. This restriction impacted approximately half
of all car trips that used to travel into what is now the LEZ Centro
{[}41{]}.

For this case study, we use spatial availability to quantify access to
opportunities by different modes in Madrid. Particularly, we demonstrate
how \(V_i^m\) can be used to derive insights into how the restriction of
car mobility in areas around/within the LEZ Centro may have allowed more
sustainable (but often slower or more costly modes) to become more
competitive.
